---
title: "E2 assignment"
output: 
   html_document:
     code_folding: hide
date: "2023-11-18"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = T, message = T)
```

# Problem 8

Performance measures In this exercise set, you will train probabilistic
classifiers which estimate $\hat{p}(y | x)$): the probability of class 𝑦
given the covariate vector x. Two commonly used performance measures for
probabilistic classifiers are accuracy and perplexity. We use the
following notation:

• $y_i \in \{0,1\}$: the true class of point 𝑖.

• $\hat{p} = \hat{p}(y = 1 | x)$: the estimated probability for the 𝑖th
point in a dataset of size 𝑛 being spam.

• $\hat{y}$: the predicted class for point i which is $\hat{y} = 1$ if
$\hat{p} \geq 0.5$ and $\hat{y} = 0$.We define the accuracy on a dataset
of n items as follows:

$$
accuracy = \Sigma_{i=1}^{n}I(y = \hat{y_i})/n
$$

$$
perplexity = \exp(-\Sigma_{i=1}^{n}log\hat{p}(y = y_i|x_i)/n)
$$

Perplexity is a transformation of the likelihood (perplexity =
exp(−loglikelihood/𝑛)), which may be the most commonly used performance
measure on probabilistic classifiers. Example values are perplexity = 1
for a perfect classifier, which always predicts the probability of one
to an actual class, and perplexity = 2 for coin flipping, which has a
predicted class probabilit $\hat{p} = 1/2$.

## Problem 8 Task a

### Question

Using one-hot encoding for $y_i$, train a logistic regression model
without Lasso or Ridge regularisation on the training data. Then: (i)
Report the model coeﬀicients. (ii) Compute and report the accuracy and
perplexity on the training and testing data. Make sure that you use no
regularisation. Notice that you may get warnings about convergence; why?
(iii) Write down the equation for the predicted class probability, given
the model coeﬀicients $\beta$ and the covariate vector $x$.

### My answer

```{r}

library(tidyverse)
library(here)

#read data
train.set <- 
  read_csv(file.path(here(), "Exercise Sets-20231117", "E2", "data_E2", "spam_train.csv"))

test.set <- 
  read_csv(file.path(here(), "Exercise Sets-20231117", "E2", "data_E2", "spam_test.csv"))
```

**The table below shows the coefficients for the model without
regularization.**

```{r}
train.set |> map(is.na) |> map(sum)
test.set |> map(is.na) |> map(sum)
```

```{r waning = T, message=T}
#load packages
library(broom)
library(kableExtra)
library(glmnet)
#train model on train set
fit <- glm(SPAM ~ ., data = train.set, family = binomial(link ="logit"))

#check output (model coefficients)
coef(fit) |> 
  tidy() |> 
  rename(
    variable = names,
    coefficient = x) |>
  mutate_if(is.numeric, ~round(.x, 3)) |> 
  DT::datatable()
#caption = "Fig P1_Task_a_1. Model coefficient on train set"

```

**The table below shows the accuracy on training and testing set.**

```{r}
# function to derive accuracy

get_accuracy <- 
  function(fit.train, test.set, true.y){
    
    # Train accuracy 
  
    ##get sample size
    n_train <-  nrow(fit.train$model[1]) #1 is list of true y 
    ##get true y
    y_train <- fit.train$model[1]
    ##get predicted probability
    prob_train <- fit.train$fitted.values
    ## assign probability to class(1/0)
    class_train <- ifelse(prob_train>=0.5, 1, 0)
    ## indicator function (0/1) for test
    indicator_train <- ifelse(class_train == y_train, 1, 0)
    ##accuracy
    accuracy_train = sum(indicator_train/n_train)
    
    # Test accuracy
    
    ## get sample size for test set
    n_test <- nrow(test.set)
    ## get true y for test set
    y_test <- test.set[,true.y]
    ## get predicted probability 
    prob_test <- predict(fit.train, test.set, type = "response")
    ## assign probability to class(1/0)
    class_test <- if_else(prob_test>=0.5, 1, 0)
    ## indicator function (0/1) for test
    indicator_test <- if_else(class_test == y_test, 1, 0)
    ## accuracy for test
    accuracy_test <- sum(indicator_test/n_test)
    
    ##print results
    output <- data.frame(accuracy_train = accuracy_train, accuracy_test =accuracy_test)
    print(output)
  }

get_accuracy(fit, test.set, "SPAM") |> 
  DT::datatable()
baseline_test_accuracy<- get_accuracy(fit, test.set, "SPAM")[1,2]
```

**The table below shows the perplexity on training and testing set.**

```{r}
# function to derive perplexity

get_perplexity <-  #true.y is the variable name of y in test.set
  function(fit.train, test.set, true.y){
    
    # Train perplexity 
  
    ##get sample size
    n_train <-  nrow(fit.train$model[1]) #1 is list of true y 
    ##get true y
    y_train <- fit.train$model[1]
    ##get predicted probability
    prob_train <- fit.train$fitted.values
    ##Indicator function (0/1)
    indicator_train <- if_else(prob_train>=0.5, 1, 0)
    ##log likelihood for train
    log_likelihood_train <- sum(y_train*log(prob_train) + (1-y_train)*log(1-prob_train))
    ##perplexity for train
    perplexity_train <- exp(-log_likelihood_train/n_train)
    
    
    # Test perplexity

    ## get sample size for test set
    n_test <- nrow(test.set)
    ## get true y for test set
    y_test <- test.set[, true.y] 
    ## get predicted probability 
    prob_test <- predict(fit.train, test.set, type = "response")
    ## og likelihood for test
     log_likelihood_test <- sum(y_test*log(prob_test) + (1-y_test)*log(1-prob_test))
    ## perplexity for test
    perplexity_test<- exp(-log_likelihood_test/n_test)
    
    ##print results
    output <- data.frame(perplexity_train = perplexity_train, perplexity_test =perplexity_test)
    print(output)
  }

# generate the table 
p1.task.a.table.1 <- get_perplexity(fit, test.set, "SPAM") |>
  mutate_if(is.numeric, ~round(.x,3))
p1.task.a.table.1|> 
  DT::datatable()

#get values to report
baseline_test_perplexity <- get_perplexity(fit, test.set, "SPAM")[1,2]

```

**Why don't converge?** The warning of convergence can be the result of
overly complex model, multi-colinearity and too few observations in
terms of predictors. In our case, we have fairly small number of
predictors against observations. I suppose the problem is mostly due to
multi-colinearity. This can be further consolidated by the common fact
that a spam email usually has more than one of these identifying
features. To tell the truth, I say it mostly because task b is about
regularization. Couldn't find a safer guess. :)

**Below is the equation for the predicted class probability, given the
model coeﬀicients** $\beta$ and the covariate vector $x$.

$$
\hat{p}(y = 1 \mid x) = \frac{1}{1 + e^{-(-55.0 + 17.5 \times x_1 + 0 \times x_2+ 17.9 \times x_3 + 34.9 \times x_4 + 38.1 \times x_5)}} \\
$$ Where $x_1$ is MISSING_FROM; $x_2$ is FROM_ADDR_WS; $x_3$ is
TVD_SPACE_RATIO; $x_4$ is LOTS_OF_MONEY; $x_5$ is
T_FILL_THIS_FORM_SHORT.

## Problem 8 Task b

### Question

Train a logistic regression model with Lasso regularisation. Find a
regularisation coeﬀicient value that performs better than the
unregularised version on the test data and has some regression
coeﬀicients equal to zero. You can do this by trying various values;
there is no need to be more sophisticated here. Report your parameters,
the regression coeﬀicients, and the accuracies and perplexities on the
testing data. Look at the predicted class probabilities for the
unregularized regressor in Task a, and your regularized regressors in
Task b. How do the distributions of the probabilities differ?

### My answer

**I trained LASSO models with a series of lambda values using train
dataset.** Then I predicted the probability of SPAM by trying different
lambda values including $\lambda$ = 0.0005, 0.05, 0.5 and 5. I selected
these $\lambda$ values arbitrarily. The table below shows their
parameters, accuracy (test set) and perplexity(test set). The first row
also exhibits un-regularized model created in task a, for the purpose of
comparing. I noticed that the lambda = 0.0005 model and un-regularized
model have same level of testing accuracy, but the former outperforms in
terms of much smaller perplexity value. Actually, the un-regularized
model's perplexity indicates it is worse than random guess. So to answer
the next part of question, I will visualize their probabilities

```{r}
# extract train set x frame
train.set.x <- 
  train.set[,which(colnames(train.set) != "SPAM")] |> 
  as.matrix()
# extract train set y vedtor
train.set.y <- 
  train.set$SPAM 

# extract test set x frame
test.set.x <- 
  test.set[,which(colnames(train.set) != "SPAM")] |> 
  as.matrix()
# extract est set y vedtor
test.set.y <- 
  test.set$SPAM 

#define lambada values to test out
grid <- 10^seq(10, -2, length = 100)

#fit a model with lasso regularization
fit.regular <- 
  glmnet(
    train.set.x, 
    train.set.y, 
    alpha = 1, #lasso
    lambda = grid,  
    family = "binomial"
    )
```

```{r}
#try various lambda values

## try lambda = 0
prediction.l0 <-  
  predict(
    fit.regular, 
    s = 0, 
    newx = test.set.x,
    type = "response"
    )

## try lambda = 0.0005
prediction.l00005 <-  
  predict(
    fit.regular, 
    s = 0.0005, 
    newx = test.set.x,
    type = "response"
    )

## try lambda = 0.05
prediction.l005 <-  
  predict(
    fit.regular, 
    s = 0.05, 
    newx = test.set.x,
    type = "response"
    )
## try lambda = 0.5
prediction.l05<-  
  predict(
    fit.regular, 
    s = 0.5, 
    newx = test.set.x,
    type = "response"
  )
## try lambda = 5
prediction.l5<-  
  predict(
    fit.regular, 
    s = 5, 
    newx= test.set.x,
    type = "response"
    )

prd.lm.s005 <-  
  predict(
    fit.regular, 
    s = 0.000026, 
    newx= test.set.x,
    type = "response"
    )
```

```{r}
# obtain accuracy and perplexity for the models created

#lambda = 0
## accuracy
predicted_classes <- ifelse(prediction.l0 > 0.5, 1, 0)
accuracy_test_l0 <- mean(predicted_classes == test.set.y)
## perplexity 
perplexity_test_l0 <- 
  exp(-mean(ifelse(test.set.y == 1, log(prediction.l0), log(1 - prediction.l0))))
## Coefficient
coef.l0 <-  
  predict(
    fit.regular, 
    s = 0, 
    newx = test.set.x,
    type = "coefficient"
    )
#lambda = 0.0005
## accuracy
predicted_classes <- ifelse(prediction.l00005 > 0.5, 1, 0)
accuracy_test_l00005 <- mean(predicted_classes == test.set.y)
## perplexity 
perplexity_test_l00005 <- 
  exp(-mean(ifelse(test.set.y == 1, log(prediction.l00005), log(1 - prediction.l00005))))
## Coefficient
coef.l00005 <-  
  predict(
    fit.regular, 
    s = 0.0005, 
    newx = test.set.x,
    type = "coefficient"
    )

#lambda = 0.05
##accuracy
predicted_classes <- ifelse(prediction.l005 > 0.5, 1, 0)
accuracy_test_l005 <- mean(predicted_classes == test.set.y)
## perplexity 
perplexity_test_l005 <- 
  exp(-mean(ifelse(test.set.y == 1, log(prediction.l005), log(1 - prediction.l005))))
## Coefficient
coef.l005 <-  
  predict(
    fit.regular, 
    s = 0.05, 
    newx = test.set.x,
    type = "coefficient"
    )
#lambda = 0.5
##accuracy
predicted_classes <- ifelse(prediction.l05 > 0.5, 1, 0)
accuracy_test_l05 <- mean(predicted_classes == test.set.y)
## perplexity 
perplexity_test_l05 <- 
  exp(-mean(ifelse(test.set.y == 1, log(prediction.l05), log(1 - prediction.l05))))
## Coefficient
coef.l05 <-  
  predict(
    fit.regular, 
    s = 0.5, 
    newx = test.set.x,
    type = "coefficient"
    )

#lambda = 5
## accuracy
predicted_classes<- ifelse(prediction.l5 > 0.5, 1, 0)
accuracy_test_l5 <- mean(predicted_classes == test.set.y)
## perplexity 
perplexity_test_l5 <- 
  exp(-mean(ifelse(test.set.y == 1, log(prediction.l5), log(1 - prediction.l5))))
## Coefficient
coef.l5 <-  
  predict(
    fit.regular, 
    s = 5, 
    newx = test.set.x,
    type = "coefficient"
    )

coefficient.l00005 <-  
  predict(
    fit.regular, 
    s = 0.0005, 
    newx = test.set.x,
    type = "coefficient"
    )
```

```{r}
p1_taskb_table <- 
  tibble(
    model = c(
      "no regularization",
      #"lambda = 0",
      "lambda = 0.0005",
      "lambda = 0.05",
      "lambda = 0.5",
      "lambda = 5"),
    `accuracy(test set)` = c(
      baseline_test_accuracy, 
      #accuracy_test_l0, 
      accuracy_test_l00005,
      accuracy_test_l005,
      accuracy_test_l05,
      accuracy_test_l5
      ), 
    `perplexity(test set)` = c(
      baseline_test_perplexity, 
      #perplexity_test_l0, 
      perplexity_test_l00005,
      perplexity_test_l005,
      perplexity_test_l05,
      perplexity_test_l5
      )
  )

p1_taskb_table



#coef.l0.parameters <- coef.l0 |> as.matrix() |> data.frame() 
coef.l00005.parameters <- coef.l00005 |> as.matrix() |> data.frame() 
coef.l005.parameters <- coef.l005 |> as.matrix() |> data.frame() 
coef.l05.parameters <- coef.l05 |> as.matrix() |> data.frame() 
coef.l5.parameters <- coef.l5 |> as.matrix() |> data.frame() 
baseline <- fit$coefficients |> data.frame()
parameters <- 
  cbind(
    baseline,
    #coef.l0.parameters,
    coef.l00005.parameters,
    coef.l005.parameters, 
    coef.l05.parameters,
    coef.l5.parameters
    ) 

colnames(parameters) <- c(
  #"l0", 
  "l00005", "l005", "l05", "l5")
parameter.table <- parameters |> t() |> data.frame()
rownames(parameter.table) <- NULL
p1_taskb_table <- cbind(p1_taskb_table, parameter.table)
```

```{r}
p1_taskb_table |> 
  mutate_if(is.numeric, ~round(.x,3)) |> 
  DT::datatable()
```

**To analyze how the distributions of probabilities differ between the
regularized and unregularized models, I visualized and overlaid their
density plots.** On each side of p\>0.5 ($\hat{y} = 1$) and
p\<0.5$\hat{y} = 0$, the LASSO model with $\lambda = 0.0005$ shows more
centered and higher peaks.

```{r}
plot(density(prediction.l00005), col = "blue", cex.main = 0.9, main = "Distribution of probability non-regularized model vs Lasso model (lambda = 0.0005)", xlab = "Predicted probability")
lines(density(fit$fitted.values), col = "red")
abline(v = 0.5, col = "black")
text(0.7, 2, "<-Classifiction cut off")
text(-0.15, 2.7, "predicted_y = 0", cex = 0.8)
text(0.65, 2.7, "predicted_y = 1", cex = 0.8)

legend("topright", c("Non-regularized model", "Lasso model"), 
       col =c("blue","red"), cex =0.7, lty=1)
```

# Problem 9

[6 points] Objective: generative Bayes classifier In this problem, you
will study the quadratic discriminant analysis (QDA) classifier.
Consider a simple case with two classes and only one feature ($K$ = 2
and $p$ = 1).

## Question Task a

Prove that the QDA classifier is not linear if the class-specific
variances differ (𝜎12 ≠ 𝜎2). Hint: This problem is from the textbook
(Problem 3, page 189). Please see the discussion in the textbook for
hints and guidance. For this problem, you should follow the arguments
laid out in Sect. 4.4.1 of the textbook, but without assuming that 𝜎12 =
𝜎2.

## My answer

In QDA, the decision boundary between two classes is determined by the
point where their respective probability distributions are equal. For a
single feature $x$ and two classes, the class-conditional densities for
a normal distribution are given by:

$$ 
p(x | y = k) = \frac{1}{\sqrt{2\pi\sigma_k^2}} \exp\left(-\frac{(x - \mu_k)^2}{2\sigma_k^2}\right) 
$$

where \$ k \$ is the class index (1 or 2), \$ \mu\_k \$ is the mean of
class \$ k \$, and \$ \sigma\_k\^2 \$ is the variance of class \$ k \$.

\\section\*{Decision Rule:}

The QDA classifier assigns a point \$ x \$ to the class for which \$ p(x
\| y = k)\$ is higher. The decision boundary is where
$$ p(x | y = 1) = p(x | y = 2) $$

Setting the class-conditional densities equal to each other, we get:

$$ \frac{1}{\sqrt{2\pi\sigma_1^2}} \exp\left(-\frac{(x - \mu_1)^2}{2\sigma_1^2}\right) = \frac{1}{\sqrt{2\pi\sigma_2^2}} \exp\left(-\frac{(x - \mu_2)^2}{2\sigma_2^2}\right) $$

Taking the logarithm of both sides simplifies this to:

$$ -\frac{(x - \mu_1)^2}{2\sigma_1^2} + \log(\sigma_2) = -\frac{(x - \mu_2)^2}{2\sigma_2^2} + \log(\sigma_1) $$

Expanding and rearranging terms, we get a quadratic equation in \$ x \$:

$$ a x^2 + b x + c = 0 $$

where \$ a \$, \$ b \$, and \$ c \$ are coefficients that depend on \$
\mu\_1 \$, \$\mu\_2 \$, \$\sigma\_1\^2 \$, and \$ \sigma\_2\^2 \$.

If \$ \sigma\_1\^2 = \sigma\_2\^2 \$, the \$ x\^2 \$ terms cancel out,
and the decision boundary becomes linear. However, if \$ \sigma\_1\^2
\neq \sigma\_2\^2 \$, the \$ x\^2 \$ term remains, making the decision
boundary quadratic, hence non-linear.

# Problem 10

Objective: naive Bayes classifier

```{r}
library(magick)
palmer <- image_read("https://const-ae.name/img/penguins_vs.png") |>
  image_convert("svg")
```

`r palmer`

In this problem, you will study the Palmer penguins by building your own
Naive Bayes classifier.

Artwork by @allison_horst

• Dataset description.

• Use penguins_train.csv as your training data and penguins_test.csv as
your testing data.

• Binary classification task: classify the penguin species as $y \in$
{Adelie, notAdelie} (not Adelie combining Gentoo and Chinstrap species)
based on four morphological and weight measurements of the individual
penguins, denoted by $x=(x_1, x_2, x_3, x_4)^T \in \mathbb{R}_4$

Naive Bayes (NB) classifier

Your task is to build your own NB classifier; you should not use a
ready-made classifier from a library. However, you do not need to create
a generic classifier (such as naive Bayes in the R e1071 library); it is
enough that your classifier works for this particular task.

The idea of NB is that the dimensions are conditionally independent,
given the class. Each class conditional feature distribution $p(x_i|y)$
is assumed to originate from an independent Gaussian distribution with
its mean $\mu$ and variance $\sigma^2$ for $i = 1,2,3,4$

## Problem 10 Task a

### Question

Compute and report each attribute's means and standard deviations separately in the training set for both classes. 

Estimate and report the class probabilities using Laplace smoothing with a pseudocount of 1 on the training set.
(You should produce a total of 18 numbers from this task.)


```{r}
# read train set
train <- 
  read_csv(file.path(here(), "Exercise Sets-20231117", "E2", "data_E2", "penguins_train.csv"))
# read test set
test <- 
  read_csv(file.path(here(), "Exercise Sets-20231117", "E2", "data_E2", "penguins_test.csv"))
```

```{r}
train$species |> table()
sum(train$species == "Adelie")
nrow(train)

```

### My answer

Continuous variable will be described at $m\pm sd$. Categorical
variables will be described as number of Adelie (percentage)

Penguins in the data have bill length of
`r mean(train$bill_length_mm) |> round(3)` $\pm$
`r sd(train$bill_length_mm)|> round(3)` $mm$; have bill depth of
`r mean(train$bill_depth_mm) |> round(3)` $\pm$
`r sd(train$bill_depth_mm)|> round(3)` $mm$; have flipper length of
`r mean(train$flipper_length_mm) |> round(3)` $\pm$
`r sd(train$flipper_length_mm)|> round(3)` $mm$; have body mass of
`r mean(train$body_mass_g) |> round(3)` $\pm$
`r sd(train$body_mass_g)|> round(3)` $g$; The number of Adelie penguins
are `r paste0(sum(train$species == "Adelie"),"(" ,sum((train$species == "Adelie")/nrow(train)*100) |> round(1), "%)")`

```{r}
# assign values 
pseudocount <- 1
n_trial <- nrow(train)
class <- table(train$species)
n_dimension <- class |> length()
# calculate 
class_probs <- (class + pseudocount) / (n_trial + n_dimension)

```



```{r}
total_count <- nrow(train)
class_counts <- table(train$species)
class_probs <- (class_counts + 1) / (total_count + length(class_counts))
class_probs[[1]]
```

Class probabilities of `r names(class)[1]` is `r class_probs[[1]] |> round(3)`; and class probabilities of `r names(class)[2]` is `r class_probs[[2]] |> round(3)`